Meta-analytic approaches and 'new' effect sizes to account for heterogeneous effects in comparative physiology

### 2021-06-07 MS/ms.Rmd
Wrote het section and also meta-regerssion section, but first draft

It will not be possible to derive an effect size that completely controls for known heterogenity at the effect size level all the time. This may because because treatment effects may not be simple linear or exponential functions of the mean response making the relationship with effect size complex [e.g., thermal performance curves; @Noble2018]. For example, dosages applied to treatments can follow quite complex non-linear relationships, with expoential or quadratic relationships in relation to some mean response. When this occurs we recommend comparative physiologists apply meta-regerssion approaches, controlling for these known treatment differences as a moderator. There are simple statisical approaches a meta-analyst can use to then make the overall effects interpretable [@Schielzeth2010]. Before diving into meta-regression models, we will first overview the meta-analytical approaches or steps commonly used.

@@

If we assume that we are interested in understanding the overall effect of, say, stress (i.e., corticosterone) on XX we might extract data from studies manipulating corticosterone on eggs. Of course, these studies are diverse, apply different dosages of corticosterone and are working across many species. If the studies are experimental in nature, than a standardised effect size, such as $ln_{RR}$ might be applicable. We then might fit an overall, multilevel meta-analytic model as follows:

$$ z_{i} = u + a_{k[i]} + sp_{k[i]} + s_{j[i]} + e_{i} + m_{i} $$

where $z_{i}$ is the *i*th effect size (*i* = 1 ... $N_{ES}$, the number of effect sizes), *u* the overall meta-analytic mean, $a_{k[i]}$ is the phylogenetic effect for species *k* applied to effect size *i*. Phylogenetic effects are assumed to be normally distributed deviates sampled from a distribution with a mean of 0 and variance $\sigma^{2}_aA$ [hereafter denoted ~ *N*$(0, \sigma^{2}_aA)$], where A is a phylogenetic correlation matrix derived from a phylogenetic tree, $sp_{k[i]}$ is the species-specific effect for the *k*th species applied to effect size *i* [assumed to be ~ *N*$(0, \sigma^{2}_{sp}I)$], $s_{j[i]}$ is the study-specific effect for the *j*th study applied to effect size *i* [assumed to be ~ *N*$(0, \sigma^{2}_{s}I)$, where I is an identity matrix], $e_{i}$ is the effect size-specific effect (or within study effect, or residuals) for the *i*th effect size [assumed to be ~ *N*$(0, \sigma^{2}_eE)$], and $m_{i}$ is the sampling variance effect around the *i*th effect as a result of variation in precision among each effect size [assumed to be  ~ *N*$(0, M)$], where M is a variance-covariance matrix with the diagonal elements being the effect size sampling variance. This model is a 'weighted' multilevel meta-analysis model because we can estimate the overall sampling variance. This is very important because it provides a way for us to understand just how much variation in effects are simplly the result of sampling differences across studies versus other, potentially improtant sources of variation such as differences in the methods applied, species differences or even real biological effects that impact the effect size a study is likely to observe. This is formalised in what we call 'heterogenetiy analysis' which we describe below. Importantly, this model provides us with the answer to our overall question: Across all the species and studies manipulating corticosterone, how much of an overall effect does corticostrone have on XX relative to control conditions?

@@

$$ I^2_{total} = \frac{\sigma^2_{total} - \sigma^2_{m}}{\sigma^2_{total}} $$
where $\sigma^2_{total}$ is the  total effect size variance (using the above multilevel model than $\sigma^2_{total} = \sigma^2_{study} + \sigma^2_{phylogeny} + \sigma^2_{species} + \sigma^2_{residual} +\sigma^2_{m}$) and $\sigma^2_{m}$ is the total sampling error variance.  Quantifying heterogenity provides a formal way to assess how much variation is the result of 'true' biological or methodological differences as opposed to simply chance (i.e., sampling variance) [@NakagawaSantos2012; @Higgins2002]. We can only conduct such analyses if we have an estimate of the sampling variance for a given effect measure because only then can we quantify how much variation is the result of sampling variability. @Senior2016 have shown that total heterogenity is often extremely high in ecological and evolutionary meta-analysis (~91%), suggesting that effect measures vary widely across and within studies. Some of the variation is clearly the result of working with different species or strains, different traits and different outcome measures, however, some of this variation also arises because of methodological differences across studies that might be related to the dosages or temperatures used for treatments. Formally quantifying and presenting heterogenity estimates from meta-analytic models can provide a way to undersand the major drivers of effect size variation by allowing researchers to produce various $I^2$ estimates which can be formally compared within and across studies. For example, we could fit the multilevel model described above to understand what proportion of variation is the result of, say, 'study' effects following @NakagawaSantos2012:

$$ I^2_{study} = \frac{\sigma^2_{study}}{\sigma^2_{study} + \sigma^2_{phylogeny} + \sigma^2_{species} + \sigma^2_{residual} +\sigma^2_{m}} $$
where $\sigma^2_{study}$ is the study specific variance, and all other variance components are for phylogeny, species, residual (effect size level) and m (total measurement/sampling error). Many readers will notice the similarities between $I^2$ and $R^2$ [@Nakgawa2013; @Nakagawa2010]. If we want to formally assess how much effect size variation is the result of methodological factors, such as temperature and dosage differences applied across studies, then we could fit our multilevel meta-regression model, including dosage or temperature (linear, quadratic or even more complex fits), and estimate how much variation is explained by these fixed effects.

** NOTE THAT WE MAY WNAT TO SWITCH POSITION, INTRODUCE METEREGRESSION FIRSTs

#### *Multi-level Metaregression to Understand Variation and Improve Interpretation*

>>>>

It will not be possible to derive an effect size that completely controls for known heterogenity at the effect size level all the time. This may be because treatment effects are not be simple linear or exponential functions of the mean response making the effect sizes we discuss above unsuitable [e.g., thermal performance curves; @Noble2018]. For example, dosages applied to treatments can follow quite complex non-linear relationships, with expoential or quadratic relationships in relation to some mean response. When this occurs we recommend comparative physiologists apply meta-regerssion approaches, controlling for these known treatment differences as a moderator. There are simple statisical approaches a meta-analyst can use to then make the overall effects interpretable [@Schielzeth2010]. Before diving into meta-regression models, we will first overview the meta-analytical approaches or steps commonly used.

@@

Assume that we are interested in understanding the overall effect of, say, stress (i.e., corticosterone - CORT) on reproduction [e.g., see @Podmola2018] we might extract data from studies manipulating corticosterone in bird eggs. Of course, these studies are diverse, apply different dosages of corticosterone and are working across many species. If the studies are experimental in nature, than a standardised effect size, such as $ln_{RR}$ or Hedges' *d* might be applicable. We then might fit an overall, multilevel meta-analytic model as follows:
$$
\begin{equation} 
Y_{i} = \beta_{0} + a_{k[i]} + sp_{k[i]} + s_{j[i]} + e_{i} + m_{i}\\
a_{k[i]}\sim N(0, \sigma^{2}_aA),\\
sp_{k[i]}\sim N(0, \sigma^{2}_{sp}I),\\
s_{j[i]}\sim N(0, \sigma^{2}_{sp}I),\\
e_{i}\sim N(0, \sigma^{2}_eI),\\
m_{i} \sim N(0, v_{i})\\
(\#eq:MLMM)
\end{equation} 
$$
where $Y_{i}$ is the *i*th effect size (*i* = 1 ... $N_{ES}$, the number of effect sizes), $\beta_{0}$ is the overall meta-analytic mean, $a_{k[i]}$ is the phylogenetic effect for species *k* applied to effect size *i*. Phylogenetic effects are assumed to be normally distributed deviates sampled from a distribution with a mean of 0 and variance $\sigma^{2}_aA$ [i.e., *N*$(0, \sigma^{2}_aA)$], where A is a phylogenetic correlation matrix derived from a phylogenetic tree, $sp_{k[i]}$ is the species-specific effect for the *k*th species applied to effect size *i*, $s_{j[i]}$ is the study-specific effect for the *j*th study applied to effect size *i*, $e_{i}$ is the effect size-specific effect (or within study effect, or residuals) for the *i*th effect size, and $m_{i}$ is the sampling variance effect around the *i*th effect as a result of variation in precision among each effect sizes, $v_{i}$, which is known. This model is a 'weighted' multilevel meta-analysis model because the estimation of *u* is weighted by the inverse sampling variance of each effect size. The fact that we can obtain an overall estimate of sampling variance (i.e., $\sigma^2_{m}$) is very important because it provides a way for us to understand just how much variation in effects are the result of sampling differences across studies versus other, potentially improtant sources of variation such as differences in the methods applied, species differences or even real biological effects that impact the effect size a study is likely to observe [@Nakagawa2017; @NakagawaSantos2012]. This is formalised in what we call 'heterogenetiy analysis' which we describe below. Importantly, this model provides us with the answer to our overall question: Across all the species and studies manipulating egg corticosterone, how much of an overall effect does corticostrone have on reproduction relative to control conditions?

Of course, the overall effect does not provide us a way to interpret its magnitude in the context of how much CORT has been applied in the treatment. The most we can say about it, is that the effect is small, moderator of large out of the sample of effect sizes extracted from this set of studies. In the next section, we show how to 'correct' the overall effect size so that its interpretation is more meaningful.

#### *Multilevel Metaregression to Understand Variation and Improve Interpretation*

Heterogenity among effects that result from known differences in how treatments are applied can also be dealt with using multilevel meta-regression models. Multilevel meta-regression models include all the same random effects that we have already discussed, but they also include fixed effects (i.e., predictors or moderators) that attempt to explain changes in the mean effect size. In otherwords, we can modify our multilevel meta-analytic model to turn it into a multilevel meta-regression model as follows:

$$ Y_{i} = \beta_{0} + \sum \beta_{j}x_{j} + a_{k[i]} + sp_{k[i]} + s_{j[i]} + e_{i} + m_{i} $$
where $\sum \beta_{j}x_{j}$ is simply the sum of all slope effects for all moderators, $x_{j}$. All other notation is the same as described above. Importantly, the variable $x_{j}$ could be any moderator collected from studies, including dosage, temperature, salinity, or pH. Some moderators will be at the level of the study (e.g., was the study experimental or observational in nature) but others could be taken at the effect size level (i.e., was the effect derived using a tempearture of 23&deg;C or 35&deg;C). By including these moderators in a meta-regression model, the interpretation of the overall meta-analytic mean, $\beta_{0}$, changes. For example, assume we collected the change in CORT dosage relative to a control group (i.e., in uL / g) for our meta-analysis on egg CORT effects on bird reproduction. If the untransformed dosage, $x_{j}$, is included in the model then the overall meta-analystic mean estimate, $\beta_{0}$, would be the mean effect of CORT relative to a control treatment, on reproduction  when the dosage difference between the treatment and control group is 0 uL /g. Of course, this interpretation does not make much sense. We can, however, apply some simple transformations to make the overall mean estimate more intuitive to understand with the added benefit that it can help improve the interpretation of non-linear parameters estimated by the model [@Schielzeth2010; @GelmanHill2007]. One simple transformation would be to centre the variable $x_{j}$, by subtracting each value from the mean to create a new input variable (i.e., $c_{i}=x_{i}-\bar{x}$, where the subscript *i* denotes each individual effect size dosage). The new variable, $c_{j}$, is now centered on the mean and can replace $x_{j}$. When the model is refit with $c_{j}$ the estimated $\beta_{0}$ can now be more intuitivly interpreted as the overall mean effect of CORT on reproduction at an average dosage difference between treatment and control groups. Importantly, by centering, we retain the original units (uL / g) of the variable, which can be very useful to allow for a more general interpretation of the effect magnitude. Of course, if one is not interested in the magnitude of effect, we could also change our question and simply model mean reproductive output as a function of dosage (often what are referred to as 'arm-based' meta-analytic models). Here, one can model the mean for each group (i.e., control and treatment group), and acount for the known sampling variance associated with each group mean. Dosage, or other continuous variables like temperature, can then be modelled (accounting for sampling variance) to understand how mean reproductive output changes. It is important to recognise that all means need to be comparable and that these models are more complex to fit. While they can be equivalent to 'contrast-based' models, they require complex interactions to be estimated [@Nakagawa2015].

Centering moderator variables, like temperature and dosage, not only provides a way to make the overall mean effect more interpretable in the face of treatment heterogenity, but it also allows more complex relationships to be fit without compromising interpretations, including non-linear parameters and interactions (e.g., polynomials). In other words, linear / main effects can still be interpreted normally in the presence of non-linear / interactions fit in the model [@Schielzeth2010; @GelmanHill2007]. This can be particularly powerful when more complex, nonlinear, dosage relationships exist. We provide examples in the online supplementary material (with R code) to show readers how teh models we describe above can be fit, and how, meta-analytic means can be adjusted for treatment heterogenity.

@@

$$ I^2_{total} = \frac{\sigma^2_{total} - \sigma^2_{m}}{\sigma^2_{total}}, \
\sigma^2_{total} = \sigma^2_{study} + \sigma^2_{phylogeny} + \sigma^2_{species} + \sigma^2_{residual} +\sigma^2_{m}
$$
where $\sigma^2_{total}$ is the  total effect size variance (using the above multilevel model) and $\sigma^2_{m}$ is the total sampling error variance.  Quantifying heterogenity provides a formal way to assess how much variation is the result of 'true' biological or methodological differences as opposed to simply chance (i.e., sampling variance) [@NakagawaSantos2012; @Higgins2002]. We can only conduct such analyses if we have an estimate of the sampling variance for a given effect measure because only then can we quantify how much variation is the result of sampling variability. @Senior2016 have shown that total heterogenity is often extremely high in ecological and evolutionary meta-analysis (~91%), suggesting that effect measures vary widely across and within studies. Some of the variation is clearly the result of working with different species or strains, different traits and different outcome measures, however, some of this variation also arises because of methodological differences across studies that might be related to the dosages or temperatures used for treatments. Formally quantifying and presenting heterogenity estimates from meta-analytic models can provide a way to undersand the major drivers of effect size variation by allowing researchers to produce various $I^2$ estimates which can be formally compared within and across studies. For example, we could fit the multilevel model described above to understand what proportion of variation is the result of, say, 'study' effects following @NakagawaSantos2012:

$$ I^2_{study} = \frac{\sigma^2_{study}}{\sigma^2_{total}} $$
where $\sigma^2_{study}$ is the study specific variance, and all other variance components are for phylogeny, species, residual (effect size level) and m (total measurement/sampling error). Many readers will notice the similarities between $I^2$ and $R^2$ [@Nakgawa2013; @Nakagawa2010]. If we want to formally assess how much effect size variation is the result of methodological factors, such as temperature and dosage differences applied across studies, then we could fit our multilevel meta-regression model, including dosage or temperature (linear, quadratic or even more complex fits), and estimate how much variation is explained by these fixed effects following @Nakgawa2013.

$$ I^2_{hetero} = \frac{\sigma^2_{fixed}}{\sigma^2_{fixed} + \sigma^2_{total}} $$

The $I^2_{hetero}$ provides a formal means to assess just how much variation in effects is teh result of known heterogeneity at the effect size level, driven by continuous moderators such as dosage, temperature, salinity, and even pH. $I^2_{hetero}$ maybe a useful measure to provide to help readers understand just how much effec variation can result from the specific treatment application.
